\pkg{cubble}: An \proglang{R} Package for Organizing and Wrangling Multivariate Spatio-temporal Data

### 2023-07-03 cubble.Rmd
section 2 proofed, small changes

Both cubble objects inherit tibble's attributes (which originates from data frames): \code{class}, \code{row.names}, and \code{names}. Additionally, both have three specialised attributes: \code{key}, \code{index}, and \code{coords}. \code{key} and \code{index} are used as introduced in the \pkg{tsibble} package [@tsibble]. In cubble, the \code{key} attribute identifies the row in the spatial cubble (given the internal use of \code{tidyr::nest()} for nesting), and when combined with the \code{index} argument, it identifies the row in the temporal cubble. Currently, cubble only supports one variable as the key, and the accepted temporal classes for \code{index} includes the base R classes \code{Date}, \code{POSIXlt}, \code{POSIXct}, as well as tsibble's \code{yearmonth}, \code{yearweek}, and \code{yearquarter} classes. The \code{coords} attribute represents an ordered pair of coordinates that can be either an unprojected pair of longitude and latitude, or a projected easting and northing value. Moreover, temporal cubbles have a special attribute called \code{spatial} to store the spatial variables. Shortcut functions are available to extract attributes from the temporal cubble object, for example, \code{spatial()} for extracting spatial variables:

@@

The spatial and temporal aspect of spatio-temporal data are often stored separately in the database. For climate data, analysts may initially receive station metadata and then query the time series based on the metadata. A (spatial) cubble object can be constructed from separate spatial and temporal tables using the function \code{make_cubble()}. The three attributes \code{key}, \code{index}, and \code{coords} need to be specified in creation. The following code creates a spatial cubble from its spatial component, \code{stations} and temporal component \code{meteo}:

@@

In \proglang{R}, there are other existing spatio-temporal data structures and this section compares and contrasts \pkg{cubble} with other existing alternatives, specifically \pkg{stars} and \pkg{sftime}. The \pkg{stars} package [@stars] uses an array structure, as oppose to tibble, to represent multivariate spatio-temporal data. While both \pkg{stars} and \pkg{cubble} support vector and raster data, it is a matter of choice on which structure to use given the application. Analysts working on satellite imageries may prefer the array structure in \pkg{stars}, while others originally working with spatio-temporal data in 2D data frames may find \pkg{cubble} easier to adopt from their existing computing workflow.

The \pkg{sftime} package [@sftime] also builds from a tibble object and its focus is on handling irregular spatio-temporal data. This means \pkg{sftime} can also handle full space-time grids and sparse space-time layouts represented in \pkg{cubble}. However, \pkg{cubble} uses nesting to avoid storing spatial variables repetitively at each timestamp. This provides memory efficiency when data is observed frequent, i.e. daily or sub-daily, or the spatial geometry is computationally expensive to store repeatedly, i.e. polygons or multipolygons. Consider the \code{climate_aus} data in the \pkg{cubble} package with `r nrow(cubble::climate_aus)` stations observed daily throughout the year 2020. In that case, the \code{sftime} object is approximately 14 times larger than the corresponding \code{cubble} object (118 MB vs. 8.5 MB).

>>>>

Both cubble objects inherit tibble's attributes (which originates from data frames): \code{class}, \code{row.names}, and \code{names}. Additionally, both have three specialised attributes: \code{key}, \code{index}, and \code{coords}, where \code{key} and \code{index} are used as introduced in the \pkg{tsibble} package [@tsibble]. In cubble, the \code{key} attribute identifies the row in the spatial cubble (given the internal use of \code{tidyr::nest()} for nesting), and when combined with the \code{index} argument, it identifies the row in the temporal cubble. Currently, cubble only supports one variable as the key. The accepted temporal classes for \code{index} includes the base R classes \code{Date}, \code{POSIXlt}, \code{POSIXct}, as well as tsibble's \code{yearmonth}, \code{yearweek}, and \code{yearquarter} classes. The \code{coords} attribute represents an ordered pair of coordinates that can be either an unprojected pair of longitude and latitude, or a projected easting and northing value. Moreover, temporal cubbles have a special attribute called \code{spatial} to store the spatial variables. Shortcut functions are available to extract attributes from the temporal cubble object, for example, \code{spatial()} for extracting spatial variables:

@@

The spatial and temporal aspect of spatio-temporal data are often stored separately in the database. For climate data, analysts may initially receive station metadata and then query the time series based on the metadata. A (spatial) cubble object can be constructed from separate spatial and temporal tables using the function \code{make_cubble()}. The three attributes \code{key}, \code{index}, and \code{coords} need to be specified. The following code creates a spatial cubble from its spatial component, \code{stations} and temporal component \code{meteo}:

@@

In \proglang{R}, there are other existing spatio-temporal data structures and this section compares and contrasts \pkg{cubble} with other existing alternatives, specifically \pkg{stars} and \pkg{sftime}. The \pkg{stars} package [@stars] uses an array structure, as opposed to a tibble, to represent multivariate spatio-temporal data. While both \pkg{stars} and \pkg{cubble} support vector and raster data, it is a matter of choice on which structure to use given the application. Analysts working on satellite imageries may prefer the array structure in \pkg{stars}, while others originally working with spatio-temporal data in 2D data frames may find \pkg{cubble} easier to adopt from their existing computing workflow.

The \pkg{sftime} package [@sftime] also builds from a tibble object and its focus is on handling irregular spatio-temporal data. This means \pkg{sftime} can also handle full space-time grids and sparse space-time layouts represented in \pkg{cubble}. However, \pkg{cubble} uses nesting to avoid storing spatial variables repetitively at each timestamp. This provides memory efficiency when data is observed frequently, i.e. daily or sub-daily, or the spatial geometry is computationally expensive to store repeatedly, i.e. polygons or multipolygons. Consider the \code{climate_aus} data in the \pkg{cubble} package with `r nrow(cubble::climate_aus)` stations observed daily throughout the year 2020. In that case, the \code{sftime} object is approximately 14 times larger than the corresponding \code{cubble} object (118 MB vs. 8.5 MB).

### 2023-07-03 cubble.tex
section 2 proofed, small changes

Both cubble objects inherit tibble's attributes (which originates from data frames): \code{class}, \code{row.names}, and \code{names}. Additionally, both have three specialised attributes: \code{key}, \code{index}, and \code{coords}. \code{key} and \code{index} are used as introduced in the \pkg{tsibble} package \citep{tsibble}. In cubble, the \code{key} attribute identifies the row in the spatial cubble (given the internal use of \code{tidyr::nest()} for nesting), and when combined with the \code{index} argument, it identifies the row in the temporal cubble. Currently, cubble only supports one variable as the key, and the accepted temporal classes for \code{index} includes the base R classes \code{Date}, \code{POSIXlt}, \code{POSIXct}, as well as tsibble's \code{yearmonth}, \code{yearweek}, and \code{yearquarter} classes. The \code{coords} attribute represents an ordered pair of coordinates that can be either an unprojected pair of longitude and latitude, or a projected easting and northing value. Moreover, temporal cubbles have a special attribute called \code{spatial} to store the spatial variables. Shortcut functions are available to extract attributes from the temporal cubble object, for example, \code{spatial()} for extracting spatial variables:

@@

In \proglang{R}, there are other existing spatio-temporal data structures and this section compares and contrasts \pkg{cubble} with other existing alternatives, specifically \pkg{stars} and \pkg{sftime}. The \pkg{stars} package \citep{stars} uses an array structure, as oppose to tibble, to represent multivariate spatio-temporal data. While both \pkg{stars} and \pkg{cubble} support vector and raster data, it is a matter of choice on which structure to use given the application. Analysts working on satellite imageries may prefer the array structure in \pkg{stars}, while others originally working with spatio-temporal data in 2D data frames may find \pkg{cubble} easier to adopt from their existing computing workflow.

The \pkg{sftime} package \citep{sftime} also builds from a tibble object and its focus is on handling irregular spatio-temporal data. This means \pkg{sftime} can also handle full space-time grids and sparse space-time layouts represented in \pkg{cubble}. However, \pkg{cubble} uses nesting to avoid storing spatial variables repetitively at each timestamp. This provides memory efficiency when data is observed frequent, i.e.~daily or sub-daily, or the spatial geometry is computationally expensive to store repeatedly, i.e.~polygons or multipolygons. Consider the \code{climate_aus} data in the \pkg{cubble} package with 639 stations observed daily throughout the year 2020. In that case, the \code{sftime} object is approximately 14 times larger than the corresponding \code{cubble} object (118 MB vs.~8.5 MB).

>>>>

Both cubble objects inherit tibble's attributes (which originates from data frames): \code{class}, \code{row.names}, and \code{names}. Additionally, both have three specialised attributes: \code{key}, \code{index}, and \code{coords}, where \code{key} and \code{index} are used as introduced in the \pkg{tsibble} package \citep{tsibble}. In cubble, the \code{key} attribute identifies the row in the spatial cubble (given the internal use of \code{tidyr::nest()} for nesting), and when combined with the \code{index} argument, it identifies the row in the temporal cubble. Currently, cubble only supports one variable as the key. The accepted temporal classes for \code{index} includes the base R classes \code{Date}, \code{POSIXlt}, \code{POSIXct}, as well as tsibble's \code{yearmonth}, \code{yearweek}, and \code{yearquarter} classes. The \code{coords} attribute represents an ordered pair of coordinates that can be either an unprojected pair of longitude and latitude, or a projected easting and northing value. Moreover, temporal cubbles have a special attribute called \code{spatial} to store the spatial variables. Shortcut functions are available to extract attributes from the temporal cubble object, for example, \code{spatial()} for extracting spatial variables:

@@

In \proglang{R}, there are other existing spatio-temporal data structures and this section compares and contrasts \pkg{cubble} with other existing alternatives, specifically \pkg{stars} and \pkg{sftime}. The \pkg{stars} package \citep{stars} uses an array structure, as opposed to a tibble, to represent multivariate spatio-temporal data. While both \pkg{stars} and \pkg{cubble} support vector and raster data, it is a matter of choice on which structure to use given the application. Analysts working on satellite imageries may prefer the array structure in \pkg{stars}, while others originally working with spatio-temporal data in 2D data frames may find \pkg{cubble} easier to adopt from their existing computing workflow.

The \pkg{sftime} package \citep{sftime} also builds from a tibble object and its focus is on handling irregular spatio-temporal data. This means \pkg{sftime} can also handle full space-time grids and sparse space-time layouts represented in \pkg{cubble}. However, \pkg{cubble} uses nesting to avoid storing spatial variables repetitively at each timestamp. This provides memory efficiency when data is observed frequently, i.e.~daily or sub-daily, or the spatial geometry is computationally expensive to store repeatedly, i.e.~polygons or multipolygons. Consider the \code{climate_aus} data in the \pkg{cubble} package with 639 stations observed daily throughout the year 2020. In that case, the \code{sftime} object is approximately 14 times larger than the corresponding \code{cubble} object (118 MB vs.~8.5 MB).
